Bayesian multi-proxy reconstruction of the early Eocene latitudinal temperature gradient

### 2023-01-15 manuscript/manuscript.Rmd
move Bayesian model description to SM

## Bayesian framework

**Model structure (Fig 2).** We model the mean temperature ($\mu$) at location $j$ as a function of absolute latitude ($abs(l)$) with a logistic regression (`growth curve` or `Richard's curve`) of the form:

$$\mu_{j}  \sim N(\nu_j, \sigma), $$

$$\nu_{j} = A~+\frac{K-A}{e^{B(abs(l_j)-M)}}~, ~~~~~ j = 1,...,n,  $$
where $A$ and $K$ denote the lower and upper asymptote, respectively, $M$ specifies the latitude of maximal growth, the most quickly with latitude, $B$ denotes the growth rate (Fig. 2), $\sigma$ denotes the residual standard deviation, and $n$ denotes the number of locations.

We infer $\mu_j$ from $m$ individual temperature observations $t_{i = 1,...m}$, derived from geochemical data, at location $j$ as

$$t_{i,j}  \sim N(\mu_j, \sigma_j), ~~~~~ i = 1,...,m,  $$
where $m$ is the number of observations at each location, and $\sigma_j$ is the estimated standard deviation of the temperatures at location $j$.

Similarly, $\mu_j$ is inferred for locations with ecological proxies from the associated normal temperature distributions with a given mean and standard deviation, $t_{\mu,j}$ and $t_{\sigma,j}$, as

$$t_{\mu,j}  \sim N(\mu_j,t_{\sigma,j}).  $$

This structure implies that $\mu_j$ is not fixed at the mean proxy temperature at location $j$, <!--of the temperatures $t_{i=1,...,m,~j}$ of location $j$-->, but is drawn towards the overall logistic regression curve, i.e. towards $\nu_j$. The pull towards $\nu_j$ tends to be high when $m$ is low, when the observations $t_{i=1,...,m,~j}$ are scattered, i.e. $\sigma_j$ is high, and/or when the overall standard deviation $\sigma$ is low. In practice, this has the desirable consequence that locations with few observations and large temperature differences between observations have less influence on the overall regression than well-sampled locations with consistent temperature observations.

**Priors (Fig2x).** In the Bayesian framework, priors need to be placed on the unknown parameters of a model. We placed weakly informative, conjugate inverse-gamma priors on $\sigma$ and $\sigma_{j=1,..n}$:
$$\sigma  \sim \sqrt{Inv-Gamma(\alpha+\frac{n}{2},\beta + 0.5\times(\mu_j-\nu_j))}, ~~~~~ j = 1,...,n,  $$
$$\sigma_j  \sim \sqrt{Inv-Gamma(\alpha+\frac{m}{2},\beta + 0.5\times(t_{i,j}-\mu_j))}, ~~~~~ i = 1,...,m,~~~~~j = 1,...,n. $$
We set $\alpha = \beta = 1$, allowing these priors to be quickly overwhelmed by the data as $n$ and $m$ increase, as we have little *a priori* knowledge of these parameters.

In contrast, we put informative priors on the regression coefficients $A$, $K$, $M$ and $B$, based on physical principles, and vaguely based on the modern climate system:

**A.** Predicted seawater surface temperatures are not allowed to be $<< -2^\circ C$, the freezing point of sea water. The highest prior density of $A$ is placed around $0^\circ C$, and it slowly tapers off towards higher temperatures. This shape is achieved by placing a skew-normal prior on the lower asymptote, specified as
$$A \sim SN(\xi=-3.0,\omega=12,\alpha_{SN}=30),$$
where $\xi$, $\omega$, and $\alpha_{SN}$ are the location, scale and shape parameters.

**K.** Input of solar energy decreases from the tropics to the poles. Hence, the latitudinal temperature gradient is broadly negative, i.e. temperature decreases with absolute latitude. This is achieved by setting $K \ge A$. The prior on the upper asymptote $K$ is a truncated normal distribution with the mean set to $K$ of the modern SST gradient, with a broad standard deviation:
$$K \sim TN(\mu_{TN} = 28, \sigma_{TN} = 10, \alpha_{TN} = A, \beta_{TN}= \infty)$$
The distribution is truncated to the left at $\alpha_{TN}=A$, but not truncated to the right ($\beta_{TN}$).

**M.** The steepness of the gradient is presumed to be highest in mid-latitudes; this is expressed with a normal prior on $M$ with the mean set to $M$ of the modern SST gradient:
$$M \sim N(42,10)$$
**B.** The steepness or growth rate $B$ of the gradient is constrained to be $\ge0$ and to not be exceedingly high, as oceanic and atmospheric heat transfer is bound to prevent very abrupt SST changes across latitudes. A gamma-distributed prior of the form
$$B \sim Gamma(\alpha_G=4.3,\beta_G=30)$$
was placed on $B$. The shape and rate parameters $\alpha_G$ and $\beta_G$ were chosen such that the highest prior density is at $B$ of the modern SST gradient, $0.11$.

@@

## Model validation

[Fig. 3]
<!--
- Modern data
- Sample based on Eocene localities
-->

To test whether our logistic regression model can adequately describe latitudinal temperature gradients, we applied a simplified version of the model to modern sea surface temperatures from Bio-Oracle [@...]. As these modern, annual sea surface temperatures estimates are associated with little uncertainty, there is no need for a hierarchical model structure, and we substituted temperature ($t$) for $\mu$ in eq. ? and ?. To verify that the modern gradient can be approximated with limited sampling, we resampled modern sea surface temperatures at modern latitudes corresponding the palaeolatitudes of the Eocene samples. This process was repeated 100 times, randomly chosing a longitude for each latitude in each repetition. The parameters of the non-hierarchical temperature model were estimated for each of the 100 samples, and the iterations after burn-in from the posterior of the parameters were pooled to generate the resulting median temperature gradient with 95 % credible intervals ($2.5^{th}$ to $97.5^{th}$ percentile).

>>>>

## Bayesian framework
We model the relationship between sea surface temperature and latitude with a generalised logistic regression (`growth curve` or `Richard's curve`). Specifically, we modelled the estimate of the mean temperature $\mu_{j}$ at location $j$ as  $\mu_{j}  \sim N(\nu_j, \sigma)$, with
$$\nu_{j} = A~+\frac{K-A}{e^{B(abs(l_j)-M)}}~, ~~~~~ j = 1,...,n,  $$
where $A$ and $K$ denote the lower and upper asymptote, respectively, $M$ specifies the latitude of maximal growth, the most quickly with latitude, $B$ denotes the growth rate (Fig. 2), $\sigma$ denotes the residual standard deviation, and $n$ denotes the number of locations. The hierarchical model structure implies that $\mu_{j}$ reflects the temperature proxies at location $j$, but is also affected by the fit to the overall latitudinal gradient. We placed priors on the unknown parameters $A$,$K$,$M$, and $B$ that were vaguely based on the modern, and observing general physical principles. For details on the model structure and the choice of priors, see the *Supplementary Material*.

@@

## Model validation
To test whether our logistic regression model can adequately describe latitudinal temperature gradients, we applied a simplified version of the model to modern sea surface temperatures from Bio-Oracle [@...]. As these modern, annual sea surface temperatures estimates are associated with little uncertainty, there is no need for a hierarchical model structure, and we substituted temperature ($t$) for $\mu$ in eq. ? and ?. To verify that the modern gradient can be approximated with limited sampling, we resampled modern sea surface temperatures at modern latitudes corresponding the palaeolatitudes of the Eocene samples. This process was repeated 100 times, randomly chosing a longitude for each latitude in each repetition. The parameters of the non-hierarchical temperature model were estimated for each of the 100 samples, and the iterations after burn-in from the posterior of the parameters were pooled to generate the resulting median temperature gradient with 95 % credible intervals ($2.5^{th}$ to $97.5^{th}$ percentile).
